---
title: "Homicides vs Educational Attainment"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
library(tidyr)
library(rmarkdown)
library(readr)
```

### Load data

```{r load-data}
# Set the Working Directory
setwd("Datasets")

# List all CSV Files
all_datasets <- list.files(pattern = "*.csv")

# Merge CSV Files into one Dataframe
df <- do.call("rbind", lapply(all_datasets, read_csv))

# Save the merged dataset as one CSV
write_csv(df, "merged_datasets.csv")

# Import the full dataset as a dataframe
gun_crime_data <- read_csv("merged_datasets.csv")

# Identify the character columns
char_cols <- sapply(gun_crime_data, is.character)

# Convert character columns to factors (if necessary) for easier analysis
gun_crime_data[char_cols] <- lapply(gun_crime_data[char_cols], as.factor)
```

* * *

## Part 1: Data 
#### Plan Phase of the PACE Framework

### How are the observations in the sample collected?
The Kaggle dataset is based on the FiveThirtyEight project, which acquired gun-related death information 
from the Centers for Disease Control and Prevention’s Multiple Cause of Death database. The original dataset 
only contained data from 2012 to 2014. This updated dataset includes data for additional years, from 2006 to 2011 
and from 2015 to 2020.

The observations are likely independent of one another, as they were randomly sampled. However, it is impossible 
to confirm this with 100% certainty, as there may be some unknown factors that could have influenced the 
independence of the observations.

As the data collected is based on observations rather than a controlled experiment, it is not possible to infer causation.

* * *

## Part 2: Research question 
#### Plan Phase of the PACE Framework

### Project Scope
The goal of this project is to investigate the correlation (if any) between the level of education of a homicide victim 
and their likelihood of being a victim of homicide. Specifically, we want to determine whether people with lower 
levels of education are more likely to be victims of homicide than people with higher levels of education.

* * *

## Part 3: Exploratory data analysis 
#### Analyze Phase of the PACE Framework

### Investigate the Dataset
Initial investigations of the dataset suggest a correlation between lower educational attainment and a higher 
rate of homicides. As educational attainment increases, the homicide rate appears to decrease.

There are several factors that could contribute to this being accurate:

- Educational attainment is associated with socioeconomic status. People with lower educational attainment are more likely to live in poverty, 
which is a risk factor for violence. Poverty can lead to stress, frustration, and a lack of opportunities, which can increase the likelihood 
of violence.

- Educational attainment is associated with cognitive ability. People with lower educational attainment may have lower cognitive abilities, 
which can make it more difficult for them to solve problems peacefully. They may also be more impulsive and less able to control their 
emotions, which can increase the risk of violence.

- Educational attainment is associated with social norms. People with lower educational attainment may be more likely to live in communities 
where violence is more common. They may also be more likely to be exposed to violent media, which can normalize violence and make it more 
likely that they will engage in violence themselves.

```{r eda}
# Initial investigation of key variables
table(gun_crime_data$race,
      gun_crime_data$education,
      gun_crime_data$intent)

ggplot(gun_crime_data, aes(x = intent, fill = race)) +
  geom_bar(stat = "count") +
  facet_wrap(~education) + 
  labs(title = "Gun Death Intent vs Educational Attainment",
       x = "Intent",
       y = "Count") +
  scale_fill_brewer(palette = "Set1") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.spacing = unit(1, "cm")
  )
```

### Data Manipulation and Cleaning
```{r format_clean}
# Keep columns for analysis
gun_crime_data_trimmed <- gun_crime_data %>%
  select(c("year","intent","race","education"))

# Remove all rows with irrelevant data for this analysis
gun_crime_data_trimmed <- gun_crime_data_trimmed[!grepl("Suicide", gun_crime_data_trimmed$intent), ]
gun_crime_data_trimmed <- gun_crime_data_trimmed[!grepl("Accidental", gun_crime_data_trimmed$intent), ]
gun_crime_data_trimmed <- gun_crime_data_trimmed[!grepl("Undetermined", gun_crime_data_trimmed$intent), ]

# Check for Missing Values
sum(is.na(gun_crime_data_trimmed))

# Replace missing values with the most frequent value
most_frequent_value <- levels(gun_crime_data_trimmed$intent)[which.max(table(gun_crime_data_trimmed$intent))]
gun_crime_data_trimmed <- gun_crime_data_trimmed %>%
  mutate(intent = replace_na(intent, most_frequent_value))

most_frequent_value <- levels(gun_crime_data_trimmed$race)[which.max(table(gun_crime_data_trimmed$race))]
gun_crime_data_trimmed <- gun_crime_data_trimmed %>%
  mutate(race = replace_na(race, most_frequent_value))

most_frequent_value <- levels(gun_crime_data_trimmed$education)[which.max(table(gun_crime_data_trimmed$education))]
gun_crime_data_trimmed <- gun_crime_data_trimmed %>%
  mutate(education = replace_na(education, most_frequent_value))

table(gun_crime_data_trimmed$race,
      gun_crime_data_trimmed$education,
      gun_crime_data_trimmed$intent)

ggplot(gun_crime_data_trimmed, aes(x = intent, fill = race)) +
  geom_bar(stat = "count") +
  facet_wrap(~education) + 
  labs(title = "# of Homicides vs Educational Attainment",
       x = "",
       y = "Count") +
  scale_fill_brewer(palette = "Set1") +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.spacing = unit(1, "cm"),
  )
```

### Conditions
```{r check_conditions}

```

* * *

## Part 4: Modeling/Inference 
#### Construct Phase of the PACE Framework
```{r modeling_inference}

```


* * *

## Part 5: Prediction/Conclusion
```{r predictions_conclusion}

```

### What Story does the Data Tell?
#### Execute Phase of the PACE Framework
