---
title: "Homicides vs Educational Attainment"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
library(tidyr)
library(rmarkdown)
library(readr)
library(viridis)
```

### Load data

```{r load-data}
# Set the Working Directory
setwd("Datasets")

# List all CSV Files
all_datasets <- list.files(pattern = "*.csv")

# Merge CSV Files into one Dataframe
df <- do.call("rbind", lapply(all_datasets, read_csv))

# Save the merged dataset as one CSV
write_csv(df, "merged_datasets.csv")

# Import the full dataset as a dataframe
gun_crime_data <- read_csv("merged_datasets.csv")

# Identify the character columns
char_cols <- sapply(gun_crime_data, is.character)

# Convert character columns to factors (if necessary) for easier analysis
#gun_crime_data[char_cols] <- lapply(gun_crime_data[char_cols], as.factor)
```

* * *

## Part 1: Data 
#### Plan Phase of the PACE Framework

### How are the observations in the sample collected?
The Kaggle dataset is based on the FiveThirtyEight project, which acquired gun-related death information 
from the Centers for Disease Control and Prevention’s Multiple Cause of Death database. The original dataset 
only contained data from 2012 to 2014. This updated dataset includes data for additional years, from 2006 to 2011 
and from 2015 to 2020.

The observations are likely independent of one another, as they were randomly sampled. However, it is impossible 
to confirm this with 100% certainty, as there may be some unknown factors that could have influenced the 
independence of the observations.

As the data collected is based on observations rather than a controlled experiment, it is not possible to infer causation.

* * *

## Part 2: Research question 
#### Plan Phase of the PACE Framework

### Project Scope
The goal of this project is to investigate the correlation (if any) between the level of education of a homicide victim 
and their likelihood of being a victim of homicide. Specifically, we want to determine whether people with lower 
levels of education are more likely to be victims of homicide than people with higher levels of education.

* * *

## Part 3: Exploratory data analysis 
#### Analyze Phase of the PACE Framework

### Investigate the Dataset
Initial investigations of the dataset suggest a correlation between lower educational attainment and a higher 
rate of homicides. As educational attainment increases, the homicide rate appears to decrease.

There are several factors that could contribute to this being accurate:

- Educational attainment is associated with socioeconomic status. People with lower educational attainment are more likely to live in poverty, 
which is a risk factor for violence. Poverty can lead to stress, frustration, and a lack of opportunities, which can increase the likelihood 
of violence.

- Educational attainment is associated with cognitive ability. People with lower educational attainment may have lower cognitive abilities, 
which can make it more difficult for them to solve problems peacefully. They may also be more impulsive and less able to control their 
emotions, which can increase the risk of violence.

- Educational attainment is associated with social norms. People with lower educational attainment may be more likely to live in communities 
where violence is more common. They may also be more likely to be exposed to violent media, which can normalize violence and make it more 
likely that they will engage in violence themselves.

```{r eda}
# Initial investigation of key variables
table(gun_crime_data$race,
      gun_crime_data$education,
      gun_crime_data$intent)

ggplot(gun_crime_data, aes(x = intent, fill = race)) +
  geom_bar(stat = "count") +
  facet_wrap(~education) + 
  labs(title = "Gun Death Intent vs Educational Attainment",
       x = "Intent",
       y = "Count") +
  scale_fill_brewer(palette = "Set1") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.spacing = unit(1, "cm")
  )
```

### Data Manipulation and Cleaning
```{r format_clean}
# Keep columns for analysis
gun_crime_data_trimmed <- gun_crime_data %>%
  select(c("year","intent","race","education"))

# Remove all rows with irrelevant data for this analysis
rows_to_remove <- grepl("suicide|accidental|undetermined", gun_crime_data_trimmed$intent, ignore.case = TRUE)
gun_crime_data_trimmed <- gun_crime_data_trimmed[!rows_to_remove, ]

# Remove all rows with irrelevant data for this analysis
if (FALSE) {
gun_crime_data_trimmed <- gun_crime_data_trimmed[!grepl("Suicide", gun_crime_data_trimmed$intent), ]
gun_crime_data_trimmed <- gun_crime_data_trimmed[!grepl("Accidental", gun_crime_data_trimmed$intent), ]
gun_crime_data_trimmed <- gun_crime_data_trimmed[!grepl("Undetermined", gun_crime_data_trimmed$intent), ]
}

# Check for Missing Values
sum(is.na(gun_crime_data_trimmed$education))
sum(is.na(gun_crime_data_trimmed$intent))

# Omit rows with missing values
if (FALSE) {
gun_crime_data_trimmed <- na.omit(df)
}

# Impute the missing rows with the mode.
mode <- names(which.max(table(gun_crime_data_trimmed$education)))
gun_crime_data_trimmed$education[is.na(gun_crime_data_trimmed$education)] <- mode
mode <- names(which.max(table(gun_crime_data_trimmed$intent)))
gun_crime_data_trimmed$intent[is.na(gun_crime_data_trimmed$intent)] <- mode

# Replace missing values with the most frequent value
if (FALSE) {
most_frequent_value <- levels(gun_crime_data_trimmed$intent)[which.max(table(gun_crime_data_trimmed$intent))]
gun_crime_data_trimmed <- gun_crime_data_trimmed %>%
mutate(intent = replace_na(intent, most_frequent_value))

most_frequent_value <- levels(gun_crime_data_trimmed$race)[which.max(table(gun_crime_data_trimmed$race))]
gun_crime_data_trimmed <- gun_crime_data_trimmed %>%
mutate(race = replace_na(race, most_frequent_value))

most_frequent_value <- levels(gun_crime_data_trimmed$education)[which.max(table(gun_crime_data_trimmed$education))]
gun_crime_data_trimmed <- gun_crime_data_trimmed %>%
mutate(education = replace_na(education, most_frequent_value))
}

table(gun_crime_data_trimmed$education,
      gun_crime_data_trimmed$intent)

ggplot(gun_crime_data_trimmed, aes(x = intent)) +
  geom_bar(stat = "count") +
  facet_wrap(~education) + 
  labs(title = "# of Homicides vs Educational Attainment",
       x = "",
       y = "Count") +
  scale_fill_brewer(palette = "Set1") +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.spacing = unit(1, "cm"),
  )
ggsave("Homicides_vs_Educational Attaintment.png", width=6, height=8)
```

### Conditions
#### Test Type
To analyze the data and determine if there is a correlation between the rate of homicide victimization 
and the level of educational attainment, a CHI-Square Independence Test will be used.

Conditions for the chi-square test: 
1. Independence: Sampled observations must be independent. 
‣ Random sample/assignment
Condition met; original location of data mentioned it was acquired through random sampling. 

‣ If sampling without replacement, n < 10% of population
Condition met.

‣ Each case only contributes to one cell in the table
It is possible that individuals may have been a part of more than one group, such as attaining more than
one educational group, such as HS/GED and College.

2. Sample size: Each particular scenario (i.e. cell) must have at least 5 expected cases
Condition met.

```{r check_conditions}

```

* * *

## Part 4: Modeling/Inference 
#### Construct Phase of the PACE Framework
H0 (nothing going on): Homicide victimization and educational attainment are independent. 
Homicide rates do not vary by educational status.

HA (something going on): Homicide victimization and educational attainment are dependent. 
Homicide rates do vary by educational status.
```{r modeling_inference}
contingency_table <- table(gun_crime_data_trimmed$intent, gun_crime_data_trimmed$education)
print(contingency_table)
chi_square_result <- chisq.test(contingency_table)

# Extract the test statistics
chi_square <- chi_square_result$statistic
df <- chi_square_result$parameter

# Set the desired confidence level
confidence_level <- 0.95

# Calculate the critical value
critical_value <- qchisq(1 - (1 - confidence_level) / 2, df)

# Calculate the margin of error
margin_of_error <- sqrt(chi_square / sum(chi_square_result$observed)) * critical_value

# Calculate the lower and upper bounds of the confidence interval
lower_bound <- chi_square / (1 - margin_of_error)
upper_bound <- chi_square / (1 + margin_of_error)
```


* * *

## Part 5: Prediction/Conclusion
Based on the p-value being much lower than the significance value of 0.05, it can be concluded that 
there is a correlation between the rate of homicide victimization and educational attainment. 
However, it is important to note that since this is an observational study and not an experiment, 
causation cannot be inferred.
```{r predictions_conclusion}
# Print the confidence interval
cat("Confidence Interval:", lower_bound, "-", upper_bound, "\n")

# Print the Chi Square Test Result
print(chi_square_result)
```

### What Story does the Data Tell?
#### Execute Phase of the PACE Framework
Based on the data analysis, it appears that there is a connection between the level of education 
attained and the probability of becoming a victim of homicide. While there are several other factors 
that contribute to this conclusion, for the purposes of this analysis, this information should suffice.

